---
title: "Exploratory Data Analysis - ESRU-EMOVI 2023"
subtitle: "Inequality of Opportunity Project"
author: "IOp Analysis Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load setup
source(here::here("src", "R", "00_setup.R"))
```

# 1. Introduction

This notebook performs exploratory data analysis (EDA) on the ESRU-EMOVI 2023 social mobility survey. The goal is to:

1. Understand the structure and quality of the data
2. Identify candidate variables for IOp analysis
3. Examine distributions of potential outcomes and circumstances
4. Assess missing data patterns

## 1.1 Load Data

```{r load-data}
# Load the cleaned data
df <- readRDS(get_path(config$paths$data_processed, "entrevistado_clean.rds"))

cat("Dataset dimensions:", nrow(df), "observations,", ncol(df), "variables\n")
```

# 2. Data Overview

## 2.1 Variable Summary

```{r var-summary}
# Create variable summary
var_summary <- tibble(
  variable = names(df),
  type = sapply(df, function(x) class(x)[1]),
  n_unique = sapply(df, function(x) length(unique(x[!is.na(x)]))),
  n_missing = sapply(df, function(x) sum(is.na(x))),
  pct_missing = sapply(df, function(x) round(mean(is.na(x)) * 100, 2))
) %>%
  arrange(desc(pct_missing))

# Show top variables by missingness
var_summary %>%
  head(20) %>%
  kable(caption = "Variables with most missing values") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 2.2 Variable Types Distribution

```{r var-types}
var_summary %>%
  count(type) %>%
  ggplot(aes(x = reorder(type, n), y = n, fill = type)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  labs(
    title = "Variable Types in EMOVI 2023",
    x = "Type",
    y = "Count"
  ) +
  theme_iop() +
  theme(legend.position = "none")
```

# 3. Sample Characteristics

## 3.1 Survey Design Variables

```{r survey-design}
# Look for survey design variables
design_vars <- c("factor", "estrato", "upm", "folio")
available_design <- design_vars[design_vars %in% names(df)]

if (length(available_design) > 0) {
  cat("Available survey design variables:", paste(available_design, collapse = ", "), "\n")

  # Summary of weights if available
  if ("factor" %in% names(df)) {
    cat("\nSurvey weights summary:\n")
    summary(df$factor)
  }
}
```

## 3.2 Geographic Distribution

```{r geography}
# Search for region/state variables
geo_vars <- names(df)[grepl("region|estado|ent|zona|urban|rural", names(df), ignore.case = TRUE)]

if (length(geo_vars) > 0) {
  cat("Geographic variables found:", paste(geo_vars, collapse = ", "), "\n\n")

  # Plot first geographic variable found
  geo_var <- geo_vars[1]
  df %>%
    filter(!is.na(!!sym(geo_var))) %>%
    count(!!sym(geo_var)) %>%
    ggplot(aes(x = reorder(!!sym(geo_var), n), y = n)) +
    geom_col(fill = "#0072B2") +
    coord_flip() +
    labs(
      title = paste("Distribution by", geo_var),
      x = NULL,
      y = "Count"
    ) +
    theme_iop()
} else {
  cat("No obvious geographic variables found. Check variable labels.\n")
}
```

# 4. Potential Outcome Variables

## 4.1 Search for Income Variables

```{r income-search}
# Search for income-related variables
income_vars <- names(df)[grepl("ingreso|income|decil|quintil|sueldo|salary",
                               names(df), ignore.case = TRUE)]

if (length(income_vars) > 0) {
  cat("Potential income variables:\n")
  for (v in income_vars) {
    label <- attr(df[[v]], "label")
    cat("  -", v, ":", ifelse(is.null(label), "(no label)", label), "\n")
  }
}
```

## 4.2 Search for Education Variables

```{r education-search}
# Search for education-related variables
educ_vars <- names(df)[grepl("educ|escol|school|estudi|grado",
                              names(df), ignore.case = TRUE)]

if (length(educ_vars) > 0) {
  cat("Potential education variables:\n")
  for (v in educ_vars[1:min(15, length(educ_vars))]) {
    label <- attr(df[[v]], "label")
    cat("  -", v, ":", ifelse(is.null(label), "(no label)", label), "\n")
  }
}
```

## 4.3 Outcome Distributions

```{r outcome-dist, eval=FALSE}
# NOTE: Uncomment and modify once outcome variables are identified

# Example for income decile
# ggplot(df, aes(x = income_decile)) +
#   geom_bar(fill = "#0072B2") +
#   labs(title = "Distribution of Income Deciles",
#        x = "Income Decile", y = "Count") +
#   theme_iop()

# Example for continuous income
# ggplot(df, aes(x = log(income + 1))) +
#   geom_histogram(bins = 50, fill = "#0072B2", alpha = 0.7) +
#   labs(title = "Distribution of Log Income",
#        x = "Log(Income + 1)", y = "Count") +
#   theme_iop()
```

# 5. Potential Circumstance Variables

## 5.1 Parental Background

```{r parental-search}
# Search for parental variables
parent_vars <- names(df)[grepl("padre|madre|papa|mama|parent|progenitor",
                                names(df), ignore.case = TRUE)]

if (length(parent_vars) > 0) {
  cat("Potential parental variables (first 20):\n")
  for (v in parent_vars[1:min(20, length(parent_vars))]) {
    label <- attr(df[[v]], "label")
    cat("  -", v, ":", ifelse(is.null(label), "(no label)", label), "\n")
  }
}
```

## 5.2 Variables at Age 14

```{r age14-search}
# Search for retrospective variables (at age 14)
age14_vars <- names(df)[grepl("14|catorce|cuando.*nino|childhood",
                               names(df), ignore.case = TRUE)]

if (length(age14_vars) > 0) {
  cat("Potential 'at age 14' variables:\n")
  for (v in age14_vars[1:min(20, length(age14_vars))]) {
    label <- attr(df[[v]], "label")
    cat("  -", v, ":", ifelse(is.null(label), "(no label)", label), "\n")
  }
}
```

## 5.3 Demographic Variables

```{r demographic-search}
# Search for demographic variables
demo_vars <- names(df)[grepl("sexo|sex|genero|edad|age|nac|birth|etni|indigen|lengua|piel|skin",
                              names(df), ignore.case = TRUE)]

if (length(demo_vars) > 0) {
  cat("Potential demographic variables:\n")
  for (v in demo_vars[1:min(20, length(demo_vars))]) {
    label <- attr(df[[v]], "label")
    cat("  -", v, ":", ifelse(is.null(label), "(no label)", label), "\n")
  }
}
```

# 6. Missing Data Analysis

## 6.1 Overall Missingness

```{r missingness-overview}
# Missingness by variable
missing_summary <- var_summary %>%
  mutate(
    missing_category = case_when(
      pct_missing == 0 ~ "No missing",
      pct_missing < 5 ~ "< 5%",
      pct_missing < 20 ~ "5-20%",
      TRUE ~ "> 20%"
    )
  )

missing_summary %>%
  count(missing_category) %>%
  ggplot(aes(x = missing_category, y = n, fill = missing_category)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_manual(values = c("No missing" = "#009E73", "< 5%" = "#56B4E9",
                                "5-20%" = "#E69F00", "> 20%" = "#D55E00")) +
  labs(
    title = "Variables by Missingness Level",
    x = "Missingness Category",
    y = "Number of Variables"
  ) +
  theme_iop() +
  theme(legend.position = "none")
```

## 6.2 High-Missing Variables

```{r high-missing}
# Variables with >20% missing
high_missing <- var_summary %>%
  filter(pct_missing > 20) %>%
  arrange(desc(pct_missing))

if (nrow(high_missing) > 0) {
  high_missing %>%
    head(20) %>%
    ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
    geom_col(fill = "#D55E00") +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
    coord_flip() +
    labs(
      title = "Variables with >20% Missing",
      x = NULL,
      y = "Percent Missing"
    ) +
    theme_iop()
}
```

# 7. Correlations Among Key Variables

```{r correlations, eval=FALSE}
# NOTE: Uncomment once key variables are identified

# Example correlation analysis
# key_vars <- c("income_decile", "education_years", "father_education",
#               "mother_education", "wealth_index")
#
# df %>%
#   select(all_of(key_vars)) %>%
#   cor(use = "pairwise.complete.obs") %>%
#   corrplot::corrplot(method = "color", type = "upper",
#                      addCoef.col = "black", tl.col = "black")
```

# 8. Data Dictionary Export

```{r export-dictionary}
# Export variable inventory for manual classification
inventory <- tibble(
  variable = names(df),
  label = sapply(names(df), function(v) {
    lbl <- attr(df[[v]], "label")
    if (is.null(lbl)) NA_character_ else lbl
  }),
  type = sapply(df, function(x) class(x)[1]),
  n_unique = sapply(df, function(x) length(unique(x[!is.na(x)]))),
  pct_missing = sapply(df, function(x) round(mean(is.na(x)) * 100, 2)),
  # Add columns for manual classification
  role = NA_character_,  # outcome, circumstance, effort, control, exclude
  notes = NA_character_
)

# Save for manual editing
write_csv(inventory, get_path(config$paths$tables, "variable_classification_template.csv"))
cat("Saved variable classification template to tables folder\n")
```

# 9. Summary and Next Steps

## 9.1 Key Findings

Based on this EDA:

1. **Sample size**: `r nrow(df)` observations
2. **Variables**: `r ncol(df)` total variables
3. **Missing data**: Varies by variable (see analysis above)

## 9.2 Action Items

1. [ ] Review variable labels and data dictionary
2. [ ] Identify specific variable names for:
   - Outcome (income, education, wealth)
   - Circumstances (parental, demographic, at-age-14)
3. [ ] Update `config/variable_roles.yaml` with actual variable names
4. [ ] Decide on missing data strategy
5. [ ] Proceed to preprocessing (`02_preprocess.R`)

## 9.3 Session Info

```{r session-info}
sessionInfo()
```
