---
title: "IOp Modeling Pipeline - ESRU-EMOVI 2023"
subtitle: "Conditional Inference Trees and Forests"
author: "IOp Analysis Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8,
  fig.align = "center",
  cache = TRUE
)

# Load setup and scripts
source(here::here("src", "R", "00_setup.R"))
source(here::here("src", "R", "utils", "iop_functions.R"))
```

# 1. Introduction

This notebook implements the full IOp modeling pipeline using:

1. **Conditional Inference Trees (ctree)** - For creating interpretable type partitions
2. **Conditional Inference Forests (cforest)** - For robust variable importance
3. **IOp metrics** - Gini-based, MLD-based, and R² measures

## 1.1 Load Data

```{r load-data}
# Load preprocessed data
df <- readRDS(get_path(config$paths$data_processed, "entrevistado_clean.rds"))

cat("Loaded data:", nrow(df), "observations,", ncol(df), "variables\n")
```

## 1.2 Define Variables

**IMPORTANT**: Update these variable names based on your EDA and the actual EMOVI codebook.

```{r define-vars}
# ============================================================================
# UPDATE THESE VARIABLE NAMES WITH ACTUAL EMOVI 2023 VARIABLES
# ============================================================================

# Outcome variable
OUTCOME_VAR <- "income_decile"  # PLACEHOLDER - update with actual var

# Circumstance variables
CIRCUMSTANCE_VARS <- c(
  # Parental background
  "father_education",    # PLACEHOLDER
  "mother_education",    # PLACEHOLDER
  "father_occupation",   # PLACEHOLDER

  # Demographics
  "sex",                 # PLACEHOLDER
  "ethnicity",           # PLACEHOLDER
  "skin_tone",           # PLACEHOLDER
  "birth_region",        # PLACEHOLDER
  "birth_cohort",        # PLACEHOLDER

  # Household at 14
  "n_books_14"           # PLACEHOLDER
)

# Survey weight
WEIGHT_VAR <- "factor"  # PLACEHOLDER

# Print status
cat("Configuration:\n")
cat("  Outcome:", OUTCOME_VAR, "\n")
cat("  Circumstances:", length(CIRCUMSTANCE_VARS), "variables\n")
cat("  Weight:", WEIGHT_VAR, "\n")
```

## 1.3 Prepare Analysis Data

```{r prepare-data, eval=FALSE}
# NOTE: This chunk will fail until variable names are updated
# Uncomment and run once variables are defined

# Check if variables exist
# all_vars <- c(OUTCOME_VAR, CIRCUMSTANCE_VARS)
# missing_vars <- all_vars[!all_vars %in% names(df)]
#
# if (length(missing_vars) > 0) {
#   stop("Missing variables: ", paste(missing_vars, collapse = ", "),
#        "\n\nPlease update variable names in the 'Define Variables' chunk above.")
# }
#
# # Prepare analysis dataset
# df_analysis <- df %>%
#   select(all_of(c(OUTCOME_VAR, CIRCUMSTANCE_VARS, WEIGHT_VAR))) %>%
#   drop_na(all_of(OUTCOME_VAR))
#
# cat("Analysis dataset:", nrow(df_analysis), "observations\n")
# cat("Complete cases:", sum(complete.cases(df_analysis)), "\n")
```

# 2. Conditional Inference Tree (ctree)

## 2.1 Model Formula

```{r ctree-formula, eval=FALSE}
# Create formula
# formula_str <- paste(OUTCOME_VAR, "~", paste(CIRCUMSTANCE_VARS, collapse = " + "))
# ctree_formula <- as.formula(formula_str)
# cat("Formula:", formula_str, "\n")
```

## 2.2 Hyperparameter Tuning

```{r ctree-tuning, eval=FALSE}
# Parameter grid
# param_grid <- expand.grid(
#   mincriterion = c(0.90, 0.95, 0.99),
#   minsplit = c(100),
#   minbucket = c(50),
#   maxdepth = c(4, 6, 8)
# )
#
# cat("Testing", nrow(param_grid), "parameter combinations\n")
#
# # Cross-validation
# set.seed(config$seeds$cv)
# cv_results <- list()
#
# for (i in 1:nrow(param_grid)) {
#   params <- param_grid[i, ]
#
#   ctrl <- partykit::ctree_control(
#     mincriterion = params$mincriterion,
#     minsplit = params$minsplit,
#     minbucket = params$minbucket,
#     maxdepth = params$maxdepth
#   )
#
#   # 5-fold CV
#   folds <- caret::createFolds(df_analysis[[OUTCOME_VAR]], k = 5)
#   fold_r2 <- numeric(5)
#
#   for (fold in 1:5) {
#     train_idx <- unlist(folds[-fold])
#     test_idx <- folds[[fold]]
#
#     model <- partykit::ctree(
#       ctree_formula,
#       data = df_analysis[train_idx, ],
#       control = ctrl
#     )
#
#     preds <- predict(model, newdata = df_analysis[test_idx, ])
#     actual <- df_analysis[[OUTCOME_VAR]][test_idx]
#
#     fold_r2[fold] <- 1 - sum((actual - preds)^2) / sum((actual - mean(actual))^2)
#   }
#
#   cv_results[[i]] <- data.frame(
#     mincriterion = params$mincriterion,
#     maxdepth = params$maxdepth,
#     mean_r2 = mean(fold_r2),
#     sd_r2 = sd(fold_r2)
#   )
# }
#
# cv_results_df <- bind_rows(cv_results) %>%
#   arrange(desc(mean_r2))
#
# kable(cv_results_df, caption = "Cross-Validation Results") %>%
#   kable_styling()
```

## 2.3 Fit Final ctree Model

```{r ctree-fit, eval=FALSE}
# Get best parameters
# best_params <- cv_results_df[1, ]
#
# cat("Best parameters:\n")
# cat("  mincriterion:", best_params$mincriterion, "\n")
# cat("  maxdepth:", best_params$maxdepth, "\n")
# cat("  CV R²:", round(best_params$mean_r2, 4), "\n")
#
# # Fit final model
# ctrl_final <- partykit::ctree_control(
#   mincriterion = best_params$mincriterion,
#   minsplit = 100,
#   minbucket = 50,
#   maxdepth = best_params$maxdepth
# )
#
# ctree_model <- partykit::ctree(
#   ctree_formula,
#   data = df_analysis,
#   control = ctrl_final
# )
#
# cat("\nTree structure:\n")
# print(ctree_model)
```

## 2.4 Visualize ctree

```{r ctree-plot, eval=FALSE, fig.width=14, fig.height=10}
# Plot the tree
# plot(ctree_model,
#      type = "simple",
#      inner_panel = node_inner,
#      terminal_panel = node_boxplot,
#      tp_args = list(id = FALSE, fill = "lightblue"))
```

# 3. Conditional Inference Forest (cforest)

## 3.1 Fit cforest Model

```{r cforest-fit, eval=FALSE}
# Set up control (using party package for unbiased variable importance)
# library(party)
#
# ctrl_forest <- cforest_unbiased(ntree = 500)
#
# cat("Fitting cforest with 500 trees...\n")
# tic()
#
# cforest_model <- cforest(
#   ctree_formula,
#   data = df_analysis,
#   controls = ctrl_forest
# )
#
# toc()
```

## 3.2 Variable Importance

```{r varimp, eval=FALSE}
# Compute conditional variable importance
# cat("Computing conditional variable importance...\n")
# tic()
#
# varimp_conditional <- varimp(cforest_model, conditional = TRUE)
#
# toc()
#
# # Create importance table
# varimp_df <- data.frame(
#   variable = names(varimp_conditional),
#   importance = varimp_conditional
# ) %>%
#   arrange(desc(importance))
#
# kable(varimp_df, caption = "Variable Importance (Conditional)") %>%
#   kable_styling()
```

## 3.3 Plot Variable Importance

```{r varimp-plot, eval=FALSE}
# varimp_df %>%
#   mutate(variable = fct_reorder(variable, importance)) %>%
#   ggplot(aes(x = importance, y = variable)) +
#   geom_point(size = 3, color = "#0072B2") +
#   geom_segment(aes(xend = 0, yend = variable), color = "#0072B2") +
#   geom_vline(xintercept = 0, linetype = "dashed") +
#   labs(
#     title = "Conditional Variable Importance (cforest)",
#     x = "Importance (reduction in MSE)",
#     y = NULL
#   ) +
#   theme_iop()
```

## 3.4 Bootstrap Variable Importance

```{r varimp-bootstrap, eval=FALSE}
# n_boot <- 50  # Reduced for speed; use 100 for final analysis
#
# cat("Bootstrapping variable importance (", n_boot, " replications)...\n")
# tic()
#
# boot_results <- list()
#
# for (b in 1:n_boot) {
#   if (b %% 10 == 0) cat("  Iteration", b, "\n")
#
#   set.seed(SEEDS$bootstrap[b])
#   boot_idx <- sample(1:nrow(df_analysis), replace = TRUE)
#   boot_data <- df_analysis[boot_idx, ]
#
#   boot_model <- cforest(ctree_formula, data = boot_data, controls = ctrl_forest)
#   boot_vi <- varimp(boot_model, conditional = TRUE)
#
#   boot_results[[b]] <- boot_vi
# }
#
# toc()
#
# # Combine and summarize
# boot_matrix <- do.call(rbind, boot_results)
#
# varimp_boot_summary <- data.frame(
#   variable = colnames(boot_matrix),
#   mean_importance = colMeans(boot_matrix),
#   sd_importance = apply(boot_matrix, 2, sd),
#   ci_lower = apply(boot_matrix, 2, quantile, 0.025),
#   ci_upper = apply(boot_matrix, 2, quantile, 0.975)
# ) %>%
#   arrange(desc(mean_importance))
#
# kable(varimp_boot_summary, digits = 4,
#       caption = "Variable Importance with Bootstrap CIs") %>%
#   kable_styling()
```

# 4. IOp Metrics

## 4.1 Calculate IOp Indices

```{r iop-metrics, eval=FALSE}
# Load IOp functions
# source(here::here("src", "R", "05_iop_metrics.R"))
#
# # Get predictions from ctree
# type_means <- predict(ctree_model, type = "response")
# types <- predict(ctree_model, type = "node")
# observed <- df_analysis[[OUTCOME_VAR]]
#
# # Calculate metrics
# iop_gini <- calculate_iop_share(observed, type_means, metric = "gini")
# iop_mld <- calculate_iop_share(observed, type_means, metric = "mld")
# iop_rsq <- calculate_iop_rsq(observed, type_means)
#
# # Create summary table
# iop_summary <- data.frame(
#   Metric = c("IOp Share (Gini)", "IOp Share (MLD)", "IOp Share (R²)",
#              "Total Gini", "Total MLD", "Number of Types"),
#   Value = c(iop_gini$iop_share, iop_mld$iop_share, iop_rsq,
#             iop_gini$ineq_total, iop_mld$ineq_total, length(unique(types)))
# )
#
# kable(iop_summary, digits = 4, caption = "IOp Metrics Summary") %>%
#   kable_styling()
```

## 4.2 MLD Decomposition

```{r mld-decomposition, eval=FALSE}
# mld_decomp <- decompose_mld(observed, types)
#
# cat("MLD Decomposition:\n")
# cat("  Total MLD:", round(mld_decomp$mld_total, 4), "\n")
# cat("  Between-type MLD:", round(mld_decomp$mld_between, 4), "\n")
# cat("  Within-type MLD:", round(mld_decomp$mld_within, 4), "\n")
# cat("  IOp Share:", round(mld_decomp$iop_share * 100, 1), "%\n")
```

## 4.3 Visualize IOp

```{r iop-viz, eval=FALSE}
# # IOp decomposition bar chart
# iop_plot_data <- data.frame(
#   Component = c("Between-Type\n(IOp)", "Within-Type\n(Residual)"),
#   Value = c(mld_decomp$mld_between, mld_decomp$mld_within),
#   Share = c(mld_decomp$iop_share, 1 - mld_decomp$iop_share)
# )
#
# ggplot(iop_plot_data, aes(x = "", y = Share, fill = Component)) +
#   geom_col(width = 0.5) +
#   geom_text(aes(label = paste0(round(Share * 100, 1), "%")),
#             position = position_stack(vjust = 0.5), size = 5) +
#   coord_flip() +
#   scale_fill_manual(values = c("#D55E00", "#56B4E9")) +
#   labs(
#     title = "Inequality Decomposition",
#     subtitle = "Share of total inequality due to circumstances vs. residual",
#     x = NULL, y = "Share of Total Inequality"
#   ) +
#   theme_iop() +
#   theme(axis.text.y = element_blank())
```

# 5. Sensitivity Analysis

## 5.1 Different Circumstance Sets

```{r sensitivity-sets, eval=FALSE}
# # Define alternative circumstance sets
# circumstance_sets <- list(
#   minimal = c("father_education", "mother_education", "sex"),
#   standard = CIRCUMSTANCE_VARS,
#   extended = c(CIRCUMSTANCE_VARS, "additional_var1", "additional_var2")  # Add more
# )
#
# sensitivity_results <- list()
#
# for (set_name in names(circumstance_sets)) {
#   vars <- circumstance_sets[[set_name]]
#   vars <- vars[vars %in% names(df_analysis)]  # Keep only existing vars
#
#   if (length(vars) < 2) next
#
#   formula_sens <- as.formula(paste(OUTCOME_VAR, "~", paste(vars, collapse = " + ")))
#
#   model_sens <- partykit::ctree(formula_sens, data = df_analysis,
#                                  control = ctrl_final)
#
#   preds <- predict(model_sens, type = "response")
#   iop <- calculate_iop_share(df_analysis[[OUTCOME_VAR]], preds, metric = "gini")
#
#   sensitivity_results[[set_name]] <- data.frame(
#     circumstance_set = set_name,
#     n_vars = length(vars),
#     iop_share = iop$iop_share
#   )
# }
#
# sensitivity_df <- bind_rows(sensitivity_results)
# kable(sensitivity_df, digits = 4, caption = "Sensitivity to Circumstance Set") %>%
#   kable_styling()
```

# 6. Save Results

```{r save-results, eval=FALSE}
# # Save models
# saveRDS(ctree_model, get_path(config$paths$models, "ctree_final.rds"))
# saveRDS(cforest_model, get_path(config$paths$models, "cforest_final.rds"))
#
# # Save tables
# save_table(cv_results_df, "ctree_cv_results")
# save_table(varimp_boot_summary, "varimp_bootstrap_ci")
# save_table(iop_summary, "iop_main_results")
#
# cat("All results saved successfully.\n")
```

# 7. Summary

## Key Results

**NOTE**: This section will be populated once the analysis runs with actual data.

- **IOp Share (Gini)**: [TO BE FILLED]
- **IOp Share (MLD)**: [TO BE FILLED]
- **Top 3 Circumstances**: [TO BE FILLED]

## Next Steps

1. Run XGBoost benchmark (`src/python/xgboost_benchmark.py`)
2. Generate comparison report
3. Create final visualizations (`03_results.Rmd`)

```{r session-info}
sessionInfo()
```
