---
title: "Final Results - Inequality of Opportunity Analysis"
subtitle: "ESRU-EMOVI 2023"
author: "IOp Analysis Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load setup
source(here::here("src", "R", "00_setup.R"))
```

# Executive Summary

This report presents the findings from our Inequality of Opportunity (IOp) analysis using the ESRU-EMOVI 2023 social mobility survey from Mexico.

**Key Finding**: [TO BE FILLED] percent of total income inequality can be attributed to circumstances outside individuals' control.

# 1. Main Results

## 1.1 IOp Estimates

```{r load-results}
# Load saved results
iop_results_file <- get_path(config$paths$tables, "iop_main_results.csv")

if (file.exists(iop_results_file)) {
  iop_results <- read_csv(iop_results_file)

  kable(iop_results, caption = "Inequality of Opportunity Metrics") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
} else {
  cat("IOp results not yet available. Run the modeling notebook first.\n")
}
```

## 1.2 IOp Decomposition Visualization

```{r iop-viz}
# Load and display IOp decomposition figure
iop_fig <- get_path(config$paths$figures, "iop_decomposition.png")

if (file.exists(iop_fig)) {
  knitr::include_graphics(iop_fig)
} else {
  cat("IOp decomposition figure not yet available.\n")
}
```

# 2. Circumstance Importance

## 2.1 Variable Importance Rankings

```{r varimp-table}
varimp_file <- get_path(config$paths$tables, "varimp_bootstrap_ci.csv")

if (file.exists(varimp_file)) {
  varimp <- read_csv(varimp_file)

  varimp %>%
    mutate(
      importance_pct = round(mean_importance / max(mean_importance) * 100, 1),
      ci = paste0("[", round(ci_lower, 4), ", ", round(ci_upper, 4), "]")
    ) %>%
    select(Variable = variable, `Importance (%)` = importance_pct,
           `95% CI` = ci) %>%
    kable(caption = "Circumstance Importance Rankings (cforest)") %>%
    kable_styling()
} else {
  cat("Variable importance results not yet available.\n")
}
```

## 2.2 Variable Importance Plot

```{r varimp-plot}
varimp_fig <- get_path(config$paths$figures, "varimp_cforest.png")

if (file.exists(varimp_fig)) {
  knitr::include_graphics(varimp_fig)
} else {
  cat("Variable importance plot not yet available.\n")
}
```

# 3. Method Comparison

## 3.1 ctree vs XGBoost

```{r comparison-table}
comparison_file <- get_path(config$paths$tables, "model_comparison.csv")

if (file.exists(comparison_file)) {
  comparison <- read_csv(comparison_file)

  kable(comparison, caption = "Method Comparison: ctree/cforest vs XGBoost") %>%
    kable_styling()
} else {
  cat("Comparison results not yet available.\n")
}
```

## 3.2 Variable Importance Comparison

```{r comparison-plot}
comparison_fig <- get_path(config$paths$figures, "varimp_comparison.png")

if (file.exists(comparison_fig)) {
  knitr::include_graphics(comparison_fig)
} else {
  cat("Comparison plot not yet available.\n")
}
```

# 4. Interpretability Analysis

## 4.1 Partial Dependence Plots

```{r pdp-grid}
pdp_fig <- get_path(config$paths$figures, "pdp_grid.png")

if (file.exists(pdp_fig)) {
  knitr::include_graphics(pdp_fig)
} else {
  cat("PDP grid not yet available.\n")
}
```

## 4.2 SHAP Summary (XGBoost)

```{r shap-summary}
shap_fig <- get_path(config$paths$figures, "shap_summary.png")

if (file.exists(shap_fig)) {
  knitr::include_graphics(shap_fig)
} else {
  cat("SHAP summary not yet available.\n")
}
```

# 5. Type Analysis

## 5.1 Type Distribution

```{r type-dist}
type_fig <- get_path(config$paths$figures, "type_distribution.png")

if (file.exists(type_fig)) {
  knitr::include_graphics(type_fig)
} else {
  cat("Type distribution plot not yet available.\n")
}
```

## 5.2 Ctree Visualization

```{r ctree-plot}
ctree_fig <- get_path(config$paths$figures, "ctree_final.png")

if (file.exists(ctree_fig)) {
  knitr::include_graphics(ctree_fig)
} else {
  cat("ctree plot not yet available.\n")
}
```

# 6. Sensitivity Analysis

## 6.1 Alternative Circumstance Sets

```{r sensitivity}
# Placeholder for sensitivity analysis results
cat("Sensitivity analysis results to be added.\n")
```

## 6.2 Subgroup Analysis

```{r subgroups}
# Placeholder for subgroup analysis
cat("Subgroup analysis results to be added.\n")
```

# 7. Comparison with Literature

## 7.1 Mexico IOp Estimates

| Study | Data | IOp Estimate | Method |
|-------|------|--------------|--------|
| Monroy-Gómez-Franco (2023) | EMOVI 2017 | ≥48% | ctree/cforest |
| Plassot et al. (2022) | EMOVI | ~40-50% | Machine learning |
| **This study** | EMOVI 2023 | [TBD]% | ctree/cforest + XGBoost |

## 7.2 International Comparison

- Latin America average: ~30-50% (varies by country)
- European countries: ~20-30% (generally lower)
- Mexico consistently shows higher IOp due to persistent social stratification

# 8. Policy Implications

Based on our findings, the most important circumstances driving inequality are:

1. **[TO BE FILLED]** - Explains X% of IOp
2. **[TO BE FILLED]** - Explains Y% of IOp
3. **[TO BE FILLED]** - Explains Z% of IOp

## Policy Recommendations

1. [TO BE DEVELOPED based on results]
2. [TO BE DEVELOPED based on results]
3. [TO BE DEVELOPED based on results]

# 9. Technical Notes

## 9.1 Data

- **Source**: ESRU-EMOVI 2023
- **Sample size**: [N observations]
- **Missing data treatment**: [Description]

## 9.2 Methods

- **ctree**: Conditional inference trees with permutation-based splitting
- **cforest**: Ensemble of ctrees with conditional variable importance
- **XGBoost**: Gradient boosted trees with SHAP explanations

## 9.3 Reproducibility

All code and data are available in this repository. To reproduce:

```bash
# 1. Set up environment
conda env create -f environment.yml
conda activate iop-analysis

# 2. Run R pipeline
Rscript src/R/01_load_validate.R
Rscript src/R/02_preprocess.R
# ... etc.

# 3. Run Python benchmark
python src/python/xgboost_benchmark.py
```

# Session Info

```{r session-info}
sessionInfo()
```

---

*Report generated on `r Sys.Date()`*
